A Transactional Memory System for Parallel Python

### 2014-06-06 talk/dls2014/paper/paper.tex
tweaks and implement some more feedback

\title{
}
%\subtitle{DLS'14}

@@

\authorinfo{Remigius Meier}
           {Department of Computer Science\\ ETH Zürich}
           {remi.meier@inf.ethz.ch}
\authorinfo{Armin Rigo}
           {www.pypy.org}
           {arigo@tunes.org}

@@

\emph{Atomic
blocks}~\cite{tim03,tim05} allow dynamic languages to indicate
the units of computation that must be executed atomically to ensure correct concurrent
execution.  Atomic blocks map nicely to transactional memory, which provide a convenient
mechansim to replace the GIL.
 Using transactions to enclose
multiple bytecode instructions, we can get the very same semantics as
provided by the GIL while possibly executing several transactions in
parallel. And by exposing these interpreter-level
transactions to the application, we
avoid the well-known problems caused by locks as applications use atomic blocks instead
of explicit acquire/release operations on locks.

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}.
We focuse here on
software based transactional memory (STM)
because they provide the unique opportunity to consider the issues of atomicity,
synchronization, and memory management (resp. garbage collection) together.
STM systems impose no a-priori bound on the size of an atomic block and provide therefore
a flexibility that is absent from current hardware-based TM systems~\cite{wayforward14}.
The downside of STM systems is runtime overhead~\cite{cascaval08,drago11}, which can
be substantial despite recent efforts to lower the overhead~\cite{warmhoff13,spear09}.
In this paper, we describe how we manage to lower the overhead of an
STM system so that it can serve as a viable replacement for the GIL by leveraging the
demands of a dynamic language like Python to integrate the STM system with the garbage
collector.

@@

\subsection{Global Interpreter Lock}

The GIL is a very simple synchronisation mechanism for supporting
multithreading in an interpreter. The basic guarantee is that the GIL
may only be released in between bytecode instructions. Thus, these
instructions are always executed atomically and in complete isolation
from others running in other threads. \emph{Atomicity} means that each
instruction and its effects seem to happen at one, indivisible point
in time. Other instructions never see inconsistent state of a
partially executed instruction (\emph{isolation}).

In addition to these guarantees, instructions are executed in a
sequential consistency model~\cite{lamport79}. This means that
the outcome of any execution of instructions in multiple threads is
equal to \emph{some} sequential execution of them.

\subsection{Transactional Memory}

Transactional memory (TM) is a concurrency control mechanism that
comes from database systems. Using transactions, we can group a series
of instructions performing operations on memory and make them happen
atomically and in complete isolation from other transactions.
Atomicity and isolation are basic properties of transactions.

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. By that, we can again provide a
sequentially consistent model for programming in multiple threads. We
can therefore use TM to directly replace the GIL. Instead of releasing
and acquiring the GIL between bytecode instructions, we commit and
start the transactions our instructions are running in.

\subsection{Python}

@@

\subsection{Synchronisation}

In Python, since the GIL is not directly exposed to the interpreter,
applications still need to synchronise memory accesses from multiple
threads using locks. Locks can be very hard to get
right~\cite{christopher10,victor11,shan08}.  They are non-composable,
have overhead, may deadlock, limit scalability, and add to the overall
complexity of the program logic. We think that \emph{atomic
blocks}~\cite{tim03,tim05} provide a better way for synchronisation.

Atomic blocks are composable, deadlock-free, higher-level and expose
useful atomicity and isolation guarantees to the application for a
series of instructions. This is why we think that the introduction
of atomic blocks to Python is a valuable contribution. Since atomicity
is a property of transactions, TM and atomic blocks are a natural fit.

@@

When a conflict is detected, we perform some simple contention
management that generally prefers the older transaction to the younger.

@@

When we store into our objects the references to other objects, we
need to be careful: addresses that are valid in one segment are not
valid in another, because they would break isolation by pointing to
the copy of the object in another segment.  Consequently, we would
need to adapt these addresses to the segment they live in, which would
prevent completely any page sharing.  Instead, we store pointers as
\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
segment.  A $SO$ can then be interpreted in any segment by adding to
it the start address of that segment.\footnote{An alternative model
would be to simulate threads using several independent processes in
which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the
state (file descriptors, external libraries, ...) is shared.} The
result is called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment
register $\%gs$\footnote{The other segment register $\%fs$ is
typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for
a similar purpose: thread-local data -- but a lot of it.}, which is
available to applications, to point to the current thread's segment
start address.  The translation from a $SO$ to a LA is done for us by
using the ``$\%gs\colon$'' prefix before any CPU instruction that
reads or writes memory using a $SO$ address.  This is very efficient
(we can do it on every object access) and some compilers support it
natively (e.g.\ clang).

In summary, translating a $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments.

@@

\item [{Read~Barrier:}] A major contribution to our performance
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  The described
  approach allows the read barrier to be a single, unconditional
  write into segment-local memory. We only need to mark an object
  as read by the current transaction. The mark is stored in an
  segment-local array indexed by a number derived from the $SO$
  of the object.

  Unlike other STM systems, the read barrier does not have to find the
  private version of the object -- our two-step address translation
  automatically resolves the reference to the private version on every
  access anyway\footnote{Additional benefit: The read barrier is not
    constrained to execute before the actual read -- both the compiler and
    the CPU are free to reorder them.}.

@@

  Here, we eagerly detect write-write conflicts by allowing only one
  transaction modifying an object at a time (by acquiring a write
  lock).\footnote{Eager write-write detection is not inherent to our
    approach; we may lift this restriction in the future.}  Part of making
  the object write-ready is also the privatisation of all pages the
  object belongs to as described above.

@@

Conflicts can be resolved by aborting one of the two transactions, or
in some cases by pausing one of them until the other commits.  If a
commit succeeds, we copy the changes made to the objects in the write
set so that they become visible in other segments.  Note that this
only includes objects that existed before the transaction and have
been modified by it.  It does not include all the new objects created
by the current transaction, which typically outnumber modified objects
by far in dynamic languages.  We allocate these new objects in pages
that are still shared, so that they do not need to be copied to other
segments at all.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g.\ it shares the
write barrier). The close integration helps in order to know more
about the lifetime of an object, as will be explained in the following
sections.

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *SO)
void stm_write(object_t *SO)
object_t *stm_allocate(ssize_t size_rounded)
STM_PUSH_ROOT(object_t *SO)
STM_POP_ROOT(object_t *SO)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment~0 is designated as
the \emph{sharing-segment} that is never assigned to a thread. The
sharing-segment is essentially read-only except during commit.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses bump-pointer allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

At the start of a transaction, the thread it is running in acquires a
segment. It may have to wait until another thread finishes its
transaction and releases a segment. Fairness is not guaranteed yet, as
we simply assume a fair scheduling policy in the operating system when
waiting on a condition variable.

Therefore, a thread may be assigned to different segments each time it
starts a transaction. Although, we try to assign it the same segment
again if possible.

@@

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of a thread's nursery into the old object space, which can be done
without stopping other threads. This is happens either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction. \remi{so this is where we mention how the GC-STM
integration is useful. highlight more or move to own section?}
\cfbolz{definitely its own section listing all the benefits}

The \textbf{old generation}, where objects are considered to be
\emph{old} and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space. Furthermore,
we optimistically re-share pages that do not need to be private
anymore.

@@

The point of the read barrier is to add the object to the read set of
the transaction. This information is needed to detect conflicts
between transactions. In other STM systems, it also resolves an object reference to
a private copy, but since the CPU performs our address translation on
every object access efficiently, we do not need to do that in our
barrier.

@@

This area can be seen as a continuous, segment-local array of bytes
that is indexed with an object's reference ($SO$) divided by 16. So
that each object has its own read marker, all objects have a size of
at least 16 bytes. Otherwise there would be false conflicts when
reading from two adjacent objects that share a single marker.
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

Note that we have three kinds of objects: \emph{young, overflow} and
\emph{old} objects. Young and overflow objects were created in the
same transaction; old objects always come from previous transactions.
The close integration between STM and the GC allows us to use this
information effectively: young objects do not trigger the slow path at
all, overflow objects only need to be handled by the GC part of the
write barrier once between collections, and only the old objects need
to be handled by the TM part once per transaction. This keeps the
overhead of the write barrier very low.

@@

For other synchronisation requirements, for example:
\begin{itemize}[noitemsep]
\item waiting for a segment to be released,
\item waiting for a transaction to abort or commit,
\item waiting for all threads to reach their safe points,
\end{itemize}
we use a set of condition variables to wait or signal other threads.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to~\cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

We evaluate our system in a Python interpreter called
PyPy\footnote{www.pypy.org} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed,
as it provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we compare between normal PyPy
using a GIL and a PyPy with STM, as well as to CPython.
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For each point in the plots, we took 5 measurements and report the
average and the standard deviation as error bars. For the JIT
benchmarks, we first let warm up the JIT by doing a few
additional iterations to get a single measurement.

@@

For the following sections we use a set of six small benchmarks
available on \cite{pypybenchs}:

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

@@

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, in order to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
\end{figure}

@@

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent~\cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary atomic blocks, let
us to believe that STM is the right choice at the moment.

@@

From the other end, there also have been attempts at lowering the
overhead of STM~\cite{warmhoff13,spear09} especially for a low number
of CPU cores. \cite{spear09} works well for some workloads, in the
case of a dynamic language VM, we have to assume any possible
workload.  \cite{warmhoff13} uses a compile-time approach to select
from multiple optimised STMs. Their partitioned FastLane STM seems to
provide good performance already on two threads using master and
helper threads. It would be interesting to see if its design also
performs well in our use case.

Similarly to us, \cite{martin09} use memory management techniques to
provide isolation between transactions. They focus on providing strong
atomicity between transactional and non-transactional code.  We,
however, use inevitable transactions to provide strong atomicity.
Additionally, while they use page fault handlers to detect conflicts,
we use traditional write barriers. Again, there are only some
superficial similarities between our approach and theirs.

@@

\bibitem{cpython} CPython. \url{www.python.org}
\bibitem{webjython} The Jython Project, \url{www.jython.org}
\bibitem{ironpython} IronPython. \url{www.ironpython.net}

>>>>

\title{A Transactional Memory System for Parallel Python}
%\subtitle{DLS'14}

@@

\authorinfo{Remigius Meier}
           {Department of Computer Science\\ ETH Zürich}
           {remi.meier@inf.ethz.ch}
\authorinfo{Armin Rigo}
           {www.pypy.org}
           {arigo@tunes.org}
\authorinfo{Carl Friedrich Bolz}
           {King's College London}
           {http://cfbolz.de}
\authorinfo{Thomas Gross}
           {Department of Computer Science\\ ETH Zürich}
           {thomas.gross@inf.ethz.ch}

@@

\emph{Atomic
blocks}~\cite{tim03,tim05} allow dynamic languages to indicate
the units of computation that must be executed atomically to ensure correct concurrent
execution.  Atomic blocks map nicely to transactional memory, which provide a convenient
mechansim to replace the GIL.
 Using transactions to enclose
multiple bytecode instructions, we can get the very same semantics as
provided by the GIL while possibly executing several transactions in
parallel. And by exposing these interpreter-level
transactions to the application, we
avoid the well-known problems caused by locks as applications use atomic blocks instead
of explicit acquire/release operations on locks~\cite{christopher10,victor11,shan08}.

There have been several attempts at replacing the GIL with
TM~\cite{nicholas06,odaira14,fuad10}.  We focus here on software based
transactional memory (STM) systems,
because they provide the unique opportunity to consider the issues of atomicity,
synchronization, and memory management (resp. garbage collection) together.
STM systems impose no a-priori bound on the size of an atomic block and provide therefore
a flexibility that is absent from current hardware-based TM systems~\cite{wayforward14}.
The downside of STM systems is runtime overhead~\cite{cascaval08,drago11}, which can
be substantial despite recent efforts to lower the overhead~\cite{warmhoff13,spear09}.
In this paper, we describe how we manage to lower the overhead of an
STM system so that it can serve as a viable replacement for the GIL by leveraging the
demands of a dynamic language like Python to integrate the STM system with the garbage
collector.

@@

% \subsection{Global Interpreter Lock}

% The GIL is a very simple synchronisation mechanism for supporting
% multithreading in an interpreter. The basic guarantee is that the GIL
% may only be released in between bytecode instructions. Thus, these
% instructions are always executed atomically and in complete isolation
% from others running in other threads. \emph{Atomicity} means that each
% instruction and its effects seem to happen at one, indivisible point
% in time. Other instructions never see inconsistent state of a
% partially executed instruction (\emph{isolation}).

% In addition to these guarantees, instructions are executed in a
% sequential consistency model~\cite{lamport79}. This means that
% the outcome of any execution of instructions in multiple threads is
% equal to \emph{some} sequential execution of them.

% \subsection{Transactional Memory}

% Transactional memory (TM) is a concurrency control mechanism that
% comes from database systems. Using transactions, we can group a series
% of instructions performing operations on memory and make them happen
% atomically and in complete isolation from other transactions.
% Atomicity and isolation are basic properties of transactions.

% If we start multiple such transactions in multiple threads, the TM
% system guarantees that the outcome of running the transactions is
% \emph{serialisable}. Meaning, the outcome is equal to some sequential
% execution of these transactions. By that, we can again provide a
% sequentially consistent model for programming in multiple threads. We
% can therefore use TM to directly replace the GIL. Instead of releasing
% and acquiring the GIL between bytecode instructions, we commit and
% start the transactions our instructions are running in.

\subsection{Python}
\remi{possibly adapt to application view \& benefits (atomic blocks)}
\remi{XXX: Also, make own section or integrate with introduction}

@@

% \subsection{Synchronisation}

% In Python, since the GIL is not directly exposed to the interpreter,
% applications still need to synchronise memory accesses from multiple
% threads using locks. Locks can be very hard to get
% right~\cite{christopher10,victor11,shan08}.  They are non-composable,
% have overhead, may deadlock, limit scalability, and add to the overall
% complexity of the program logic. We think that \emph{atomic
% blocks}~\cite{tim03,tim05} provide a better way for synchronisation.

% Atomic blocks are composable, deadlock-free, higher-level and expose
% useful atomicity and isolation guarantees to the application for a
% series of instructions. This is why we think that the introduction
% of atomic blocks to Python is a valuable contribution. Since atomicity
% is a property of transactions, TM and atomic blocks are a natural fit.

@@

When a conflict is detected, we perform some simple contention
management that generally prefers the older transaction to the younger.
This gives long transactions a better chance to succeed.

@@

When we store into our objects the references to other objects, we
need to be careful: addresses that are valid in one segment are not
valid in another, because they would break isolation by possibly pointing to
the copy of the object in another segment.  Consequently, we would
need to adapt these addresses to the segment they live in, which would
prevent completely any page sharing because then the binary contents
differ between segments.  Instead, we store pointers as
\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
segment.  An $SO$ can then be interpreted in any segment by adding to
it the start address of that segment.\footnote{An alternative model
would be to simulate threads using several independent processes in
which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the
state (file descriptors, external libraries, ...) is shared.} The
result is called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment
register $\%gs$\footnote{The other segment register $\%fs$ is
typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for
a similar purpose: thread-local data -- but a lot of it.}, which is
available to applications, to point to the current thread's segment
start address.  The translation from an $SO$ to an LA is done for us by
using the ``$\%gs\colon$'' prefix before any CPU instruction that
reads or writes memory using an $SO$ address.  This is very efficient
-- we can do it on every object access -- and some compilers support it
natively (e.g.\ clang).

In summary, translating an $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments.

@@

\item [{Read~Barrier:}] A major contribution to our performance
  is the simplicity of the read barrier: a large program, like an
  interpreter for a complex language, contains a number of reads that
  typically vastly exceeds the number of writes.  The described
  approach allows the read barrier to be a single, unconditional
  write into segment-local memory. We only need to \emph{mark} an object
  as read by the current transaction. The mark is stored in a
  segment-local array indexed by a number derived from the $SO$
  of the object.

  Unlike other STM systems, the read barrier does not have to find the
  private version of the object -- our two-step address translation
  automatically resolves the reference to the private version on every
  access anyway.

@@

  Here, we eagerly detect write-write conflicts by allowing only one
  transaction modifying an object at a time using write
  locks.\footnote{Eager write-write detection is not inherent to our
    approach; we may lift this restriction in the future.}  Part of making
  the object write-ready is also the privatisation of all pages the
  object belongs to, as described above.

@@

Conflicts can be resolved by aborting one of the two transactions, or
in some cases by pausing the writing one until the reader commits.  If a
commit succeeds, we copy the changes made to the objects in the write
set so that they become visible in other segments.  Note that this
only includes objects that existed before the transaction and have
been modified by it.  It does not include all the new objects created
by the current transaction, which typically outnumber modified objects
by far in dynamic languages.  We allocate these new objects in pages
that are still shared, so that they do not need to be copied to other
segments at all.

@@

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g.\ it shares the
write barrier).

@@

\begin{lstlisting}
void stm_start_transaction()
void stm_commit_transaction()
void stm_read(object_t *SO)
void stm_write(object_t *SO)
object_t *stm_allocate(size_t size_rounded)
STM_PUSH_ROOT(object_t *SO)
STM_POP_ROOT(object_t *SO)
\end{lstlisting}

\lstinline!stm_start_transaction()!  starts a transaction in the
current thread. Internally, it uses \lstinline!setjmp()! to remember
where the transaction started in case of an abort.
\lstinline!stm_commit_transaction()! tries to commit the current
transaction, aborting it otherwise. \lstinline!stm_read()!,
\lstinline!stm_write()!  perform
a read or a write barrier on the object given as the argument, and
\lstinline!stm_allocate()!  allocates a new object with the specified
size.  \lstinline!STM_PUSH_ROOT()! and \lstinline!STM_POP_ROOT()!
push and pop objects onto the shadow stack\footnote{A stack for pointers
  to GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}} of the current thread.  References to objects
have to be saved using this stack locally around calls that may cause a GC
cycle to happen, and also while there is no transaction
running. Otherwise these references may not be valid any more because
the object moved or was seen as garbage. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment~0 is designated as
the \emph{sharing-segment} that is never assigned to a thread. The
sharing-segment is essentially read-only, except during commit.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information for bookkeeping.
\item [{Read~markers:}] These are private pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This private area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses bump-pointer allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

At the start of a transaction, the thread needs to acquire a
segment. It may have to wait until another thread finishes its
transaction and releases its segment. Fairness is not guaranteed yet, as
we simply assume a fair scheduling policy in the operating system when
waiting on a condition variable.

Therefore, a thread may be assigned to different segments each time it
starts a transaction. Although, we try to assign it the same segment
again if possible to maximise the effectiveness of caches and the
page privatisation.

@@

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of a thread's nursery into the old object space, which can be done
without stopping other threads. This happens either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are \emph{old} and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction. Other and later transactions will treat them
like regular \emph{old} objects.

The \textbf{old generation}, where objects are considered to be
\emph{old} and never move again, is collected by \emph{major
  collections}.  These collections are implemented in a stop-the-world
kind of way and first force minor collections in all threads. The
major goal is to free objects in the old objects space.

Furthermore, we optimistically re-share pages during major collections.
This is another example where the STM-GC cooperation makes sense. We
regard private pages as additional memory that may be collected.
Pages should usually stay private for some time, but at some regular
intervals, we still want to try to re-share pages in case they actually
stay unmodified in the future. Major collections are allowed to do
that because all threads are synchronised, and we can re-share all
private pages that do not currently contain an object in some
transaction's  write set.

@@

The point of the read barrier is to add the object to the read set of
the transaction. This information is needed to detect conflicts
between transactions. In other STM systems, it also resolves an object reference to
a private copy, but since the CPU performs our address translation on
every object access efficiently, we can avoid this extra overhead.

@@

This area can be seen as a continuous, segment-local array of bytes
that is indexed with an object's reference ($SO$) divided by 16. So
that each object has its own read marker, all objects have a size of
at least 16 bytes. Otherwise there could be false conflicts when
reading from two adjacent objects that share a single marker.
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for
CPUs\footnote{Additional benefit: The read barrier is not constrained
to execute before the actual read -- both the compiler and the CPU are
free to reorder or group them.}:

@@

Note that we have three kinds of objects: \emph{young, overflow} and
\emph{old} objects. Young and overflow objects were created in the
same transaction; old objects always come from previous transactions.
The close integration between STM and the GC allows us to use this
information effectively: young objects do not trigger the slow path at
all, overflow objects only need to be handled by the GC part of the
write barrier once between collections, and only the old objects need
to be handled by the STM part \textbf{once per transaction}. This keeps the
overhead of the write barrier very low.

@@

% For other synchronisation requirements, for example:
% \begin{itemize}[noitemsep]
% \item waiting for a segment to be released,
% \item waiting for a transaction to abort or commit,
% \item waiting for all threads to reach their safe points,
% \end{itemize}
% we use a set of condition variables to wait or signal other threads.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
  in order to let long transactions succeed more often
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to~\cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

We evaluate our system in a Python interpreter called
PyPy~\cite{pypy} (version 2.3). PyPy is an implementation of an
interpreter for the Python language. It has a special focus on speed
and provides a just-in-time (JIT) compiler to speed up applications
running on top of it. For comparison, we compare between normal PyPy
using a GIL and a PyPy with STM, as well as to CPython.
\begin{description}
\item[CPython] (version 2.7.6) is the reference implementation of the Python
  language. It is the most widely used interpreter for this language.
  The implementation uses a GIL for synchronisation in multi-threaded
  execution and it does not feature a JIT compiler.
\end{description}

@@

For each point in the plots, we took 5 measurements and report the
average and the standard deviation as error bars. For the JIT
benchmarks, we first let the JIT warm up by doing a few
additional iterations to get a single measurement.

@@

For the following sections we use a set of six small benchmarks
available at~\cite{pypybenchs}:

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing execution time between interpreters without a JIT
    and using small input sizes\label{fig:performance-nojit}}
\end{figure}

@@

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, in order to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times (denoted with ``(large)'' in benchmark names).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing execution time between interpreters with JIT
    and large input sizes\label{fig:performance-jit}}
\end{figure}

@@

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent~\cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary-sized atomic blocks, let
us to believe that STM is the right choice at the moment.

@@

From the other end, there also have been attempts at lowering the
overhead of STM~\cite{warmhoff13,spear09} especially for a low number
of CPU cores. \cite{spear09} works well for some workloads, in the
case of a dynamic language VM, however, we have to assume any possible
workload.  \cite{warmhoff13} uses a compile-time approach to select
from multiple optimised STMs. Their partitioned FastLane STM seems to
provide good performance already on two threads using master and
helper threads. It would be interesting to see if its design also
performs well in our use case.

Similarly to us, \cite{martin09} use memory management techniques to
provide isolation between transactions. They focus on providing strong
atomicity between transactional and non-transactional code.  We,
however, use inevitable transactions to provide strong atomicity
for non-reversible operations and otherwise do not have to deal
with non-transactional code.  Additionally, while they use page fault
handlers to detect conflicts, we use traditional write barriers. There
are only some superficial similarities between our approach and
theirs.

@@

\bibitem{cpython} CPython. \url{www.python.org}
\bibitem{webjython} The Jython Project, \url{www.jython.org}
\bibitem{ironpython} IronPython. \url{www.ironpython.net}
\bibitem{pypy} PyPy Project. \url{www.pypy.org}
